---
title: "eBird Data Exploration"
author: "Badhia Yunes Katz"
date: "July 2025"
output: rmarkdown::github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.width=15, fig.height=10)
library(tidyverse)
library(httr)
library(jsonlite)
library(auk)
library(dplyr)
library(rnaturalearth)
library(sf)
library(ggplot2)
```

The purpose of this markdown is to bird data. This data was sourced from [CornellLab eBird Data and Tools Site](https://science.ebird.org/en/use-ebird-data). 

Data product citation:
Fink, D., T. Auer, A. Johnston, M. Strimas-Mackey, S. Ligocki, O. Robinson, W. Hochachka, L. Jaromczyk, C. Crowley, K. Dunham, A. Stillman, I. Davies, A. Rodewald, V. Ruiz-Gutierrez, C. Wood. 2023. eBird Status and Trends, Data Version: 2022; Released: 2023. Cornell Lab of Ornithology, Ithaca, New York. https://doi.org/10.2173/ebirdst.2022

### Overview


```{r include=FALSE}
bbox_url <- "https://api.ebird.org/v2/data/obs/geo/recent"

api_key = Sys.getenv("ebird_key")

response <- GET(
  url = bbox_url,
  query = list(
    lat = 42.0,
    lng = -121.5,
    dist = 50,
    maxResults = 500
  ),
  add_headers("X-eBirdApiToken" = api_key)
)

# Parse results
data <- fromJSON(content(response, "text", encoding = "UTF-8"))

# View species
unique_species <- unique(data$comName)
print(unique_species)
```


```{r include=FALSE}
# Setting up GIS files
gpkg_file <- "data/gis-data.gpkg"
dir.create(dirname(gpkg_file), showWarnings = FALSE, recursive = TRUE)

# political boundaries
# land border with lakes removed
ne_land <- ne_download(scale = 50, category = "cultural",
                       type = "admin_0_countries_lakes",
                       returnclass = "sf") |>
  filter(CONTINENT %in% c("North America", "South America")) |>
  st_set_precision(1e6) |>
  st_union()
# country boundaries
ne_countries <- ne_download(scale = 50, category = "cultural",
                       type = "admin_0_countries_lakes",
                       returnclass = "sf") |>
  select(country = ADMIN, country_code = ISO_A2)
# state boundaries for united states
ne_states <- ne_download(scale = 50, category = "cultural",
                       type = "admin_1_states_provinces",
                       returnclass = "sf") |> 
  filter(iso_a2 == "US") |> 
  select(state = name, state_code = iso_3166_2)
# country lines
# downloaded globally then filtered to north america with st_intersect()
ne_country_lines <- ne_download(scale = 50, category = "cultural",
                                type = "admin_0_boundary_lines_land",
                                returnclass = "sf") |> 
  st_geometry()
lines_on_land <- st_intersects(ne_country_lines, ne_land, sparse = FALSE) |>
  as.logical()
ne_country_lines <- ne_country_lines[lines_on_land]
# states, north america
ne_state_lines <- ne_download(scale = 50, category = "cultural",
                              type = "admin_1_states_provinces_lines",
                              returnclass = "sf") |>
  filter(ADM0_A3 %in% c("USA", "CAN")) |>
  mutate(iso_a2 = recode(ADM0_A3, USA = "US", CAN = "CAN")) |> 
  select(country = ADM0_NAME, country_code = iso_a2)

# save all layers to a geopackage
unlink(gpkg_file)
write_sf(ne_land, gpkg_file, "ne_land")
write_sf(ne_countries, gpkg_file, "ne_countries")
write_sf(ne_states, gpkg_file, "ne_states")
write_sf(ne_country_lines, gpkg_file, "ne_country_lines")
write_sf(ne_state_lines, gpkg_file, "ne_state_lines")
```

```{r echo=TRUE}
# downloaded region == Klamath, Oregon, United States (US), as a test

#sampling data
f_sed <- "data-raw/ebird_data/ebd_US-OR-035_smp_relMay-2025_sampling.txt"
checklists <- read_sampling(f_sed)
glimpse(checklists)

checklists_traveling <- filter(checklists, protocol_name == "Traveling")
ggplot(checklists_traveling) +
  aes(x = effort_distance_km) +
  geom_histogram(binwidth = 1, 
                 aes(y = after_stat(count / sum(count)))) +
  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
  labs(x = "Distance traveled [km]",
       y = "% of eBird checklists",
       title = "Distribution of distance traveled on eBird checklists")
```

```{r echo=FALSE}
# observation data

f_ebd <- "data-raw/ebird_data/ebd_US-OR-035_smp_relMay-2025.txt"
observations <- read_ebd(f_ebd)
glimpse(observations)
```

### Data Summaries

Table format data summaries for states derived from eBird Status and Trends data products. Below, we’ll explore how to access the summary for California and Oregon state, what data each column contains, and how to interpret the information.

## California Regional Status

  - birds included on this spreadsheet are the species with non-zero percent of population estimates within the state during at least one season of the year
  - If a species appears to be missing, it’s possible that its percent of population was low enough to round to zero, or perhaps the species was omitted from models for some other reason


```{r echo=FALSE}
ca_status <- read_csv("data-raw/ebird_data/CA_CSV/CA_regional_status_2023.csv") |> glimpse()
```

Column names

```{r echo=FALSE}
colnames(ca_status)
```

  - The first four columns include the name for the state, followed by the species code, common name, and scientific name for each species (state_name, species_code, common_name, scientific_name)
  - The next set of four columns gives the percent of the species global population which occurs in your state, broken up into four seasons.(breeding season, post-breeding migration, nonbreeding season, pre-breeding season. For resident species, such as Clark’s Nutcracker, all four values will be the same.)
  - next set of three columns:  
    - `max_week_season` - season in which the percent of a bird’s population reaches its maximum within
    - `max_week` - the week in which it occurs
    - `max_week_percent_pop` - maximum percent estimate which corresponds to the week
    
Example: 

- During the week of May 31st as the midpoint, nearly 1.9% of the global population of Osprey is present in California

```{r echo=FALSE}
ca_status |> 
  filter(common_name == "Osprey") |> 
  select(common_name, max_week_season, max_week, max_week_percent_pop) |> 
  glimpse()
```

- `state_rank_breeding`: first number is state rank, and second number is the total number of state with a non-zero breeding presence of the species. (e.g. 46 states have a non-zero breeding percent of population for Osprey, and out of these, CA has the 8th highest percent of population during breeding season)


```{r echo=FALSE}
ca_status |> 
  filter(common_name == "Osprey") |> 
  select(common_name, max_week_season, max_week, max_week_percent_pop, state_rank_breeding) |> 
  glimpse()
```

- `breeding_habitat_primary` and `breeding_habitat_secondary`: categories come from the Avian Conservation Assessment Database created and maintained by Partners in Flight



## California Regional Trends 

```{r echo=FALSE}
ca_trends <- read_csv("data-raw/ebird_data/CA_CSV/CA_regional_trends_2023.csv") |> glimpse()
```
```{r echo=FALSE}
colnames(ca_trends)
```

 
 
 ## Custom Download (Klamath, OR as example)
 
 - These data was downloaded in the [EBD eBird portal](https://ebird.org/data/download/ebd)
 
 #TODO decide if this approach is good, then add other regions
 #TODO look into clipping method
 
```{r echo=FALSE}
ebd_file <- "data-raw/ebird_data/ebd_US-OR-035_smp_relMay-2025.txt"

klamath <- read_ebd(ebd_file)

klamath_filtered <- klamath |>
  filter(all_species_reported == 1,
         protocol_name %in% c("Traveling", "Stationary"),
         effort_distance_km <= 5,
         duration_minutes <= 120)


# summarize reporting rate - adding month and year

klamath <- klamath_filtered |>
  mutate(year = year(observation_date),
         month = month(observation_date),
         ym = paste(year, sprintf("%02d", month), sep = "-"))


# calculate reporting rate per species
species_reporting <- klamath |>
  group_by(common_name, observer_id, sampling_event_identifier) |>
  summarise(present = 1, .groups = "drop") |>
  group_by(common_name) |>
  summarise(num_checklists = n(),
            .groups = "drop") |>
  arrange(desc(num_checklists))

# Total number of checklists
total_checklists <- klamath |>
  select(sampling_event_identifier) |>
  distinct() |>
  nrow()

species_reporting <- species_reporting |>
  mutate(reporting_rate = num_checklists / total_checklists)

 
# viz
species_reporting |>
  top_n(20, reporting_rate) |>
  ggplot(aes(x = reorder(common_name, reporting_rate), y = reporting_rate)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 20 Most Reported Species in Klamath County",
       y = "Reporting Rate", x = "")

```

Temporal trend analysis

```{r echo=FALSE}
# temporal trend analysis

klamath_filtered_2 <- klamath_filtered |>
  mutate(year = year(observation_date),
         month = month(observation_date),
         ym = format(observation_date, "%Y-%m"))


species_per_year <- klamath_filtered_2 |>
  group_by(common_name, year, sampling_event_identifier) |>
  summarise(present = 1, .groups = "drop") |>
  distinct(common_name, year, sampling_event_identifier) |>
  count(common_name, year, name = "n_reports")

# Merge with checklist counts
reporting_rate <- left_join(species_per_year, checklists_per_year, by = "year") |>
  mutate(reporting_rate = n_reports / n_checklists)


top_species <- reporting_rate |>
  group_by(common_name) |>
  summarise(mean_rate = mean(reporting_rate, na.rm = TRUE)) |>
  slice_max(mean_rate, n = 5) |>
  pull(common_name)

reporting_rate |>
  filter(common_name %in% top_species) |>
  ggplot(aes(x = year, y = reporting_rate, color = common_name)) +
  geom_line() +
  geom_point() +
  labs(title = "Reporting Rate Trends Over Time",
       x = "Year", y = "Reporting Rate",
       color = "Species") +
  theme_minimal()

```

Monthly reporting 

```{r echo=FALSE}
# Seasonal Trends (by Month)

reporting_by_month <- klamath_filtered_2 |>
  group_by(common_name, ym, sampling_event_identifier) |>
  summarise(present = 1, .groups = "drop") |>
  distinct(common_name, ym, sampling_event_identifier) |>
  count(common_name, ym, name = "n_reports") |>
  separate(ym, into = c("year", "month"), sep = "-") |>
  mutate(year = as.integer(year), month = as.integer(month))

# Merge with total checklists per month
checklists_by_month <- klamath_filtered_2 |>
  mutate(ym = format(observation_date, "%Y-%m")) |>
  select(ym, sampling_event_identifier) |>
  distinct() |>
  count(ym, name = "n_checklists") |>
  separate(ym, into = c("year", "month"), sep = "-") |>
  mutate(year = as.integer(year), month = as.integer(month))

monthly_reporting <- left_join(reporting_by_month, checklists_by_month, by = c("year", "month")) |>
  mutate(reporting_rate = n_reports / n_checklists) |> glimpse()

```
 
 
