---
title: "eBird Data Exploration"
author: "Badhia Yunes Katz"
date: "July 2025"
output: rmarkdown::github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.width=15, fig.height=10)
library(tidyverse)
library(httr)
library(jsonlite)
library(auk)
library(dplyr)
library(rnaturalearth)
library(sf)
library(ggplot2)
library(readr)
library(viridis)
library(lubridate)
Sys.setenv(PATH = paste(Sys.getenv("PATH"), "C:/Program Files/Git/usr/bin", sep = ";"))


api_key <- "ebird_key"
radius_km <- 80                 
sample_size <- 15     
date <- "2024/06/15"  
```

Some limitations

  - species preference (personal preferences)
  - spacial and temporal bias (weekends or holidays - May is popular month: near homes or easily accesible areas)
  - skill level of observer
  - #TODO add more langugage



```{r eval=FALSE, include=FALSE}
# using ebird API

# Load Klamath Basin polygon
klamath_basin <- st_read("data-raw/klamath_basin_outline/R8_FAC_Klamath_Basin_WFL1.shp")
klamath <- st_transform(klamath_basin, crs = 4326)

# picking sample points  
points <- st_sample(klamath, size = sample_size, type = "random") |> 
  st_sf() |> 
  mutate(point_id = row_number())

# query eBird points
query_point <- function(lat, lng, date, dist, key) {
  url <- paste0(
    "https://api.ebird.org/v2/data/obs/geo/historic/",
    date,
    "?lat=", lat,
    "&lng=", lng,
    "&dist=", dist,
    "&maxResults=500")
  
  resp <- GET(url, add_headers("X-eBirdApiToken" = key))
  if (http_status(resp)$category != "Success") return(NULL)
  
  dat <- fromJSON(content(resp, "text", encoding = "UTF-8"))
  if (length(dat) == 0) return(NULL)
  
  unique_species <- unique(dat$comName)
  return(unique_species)
}

# richness 
results <- lapply(seq_len(nrow(points)), function(i) {
  coords <- st_coordinates(points[i, ])
  species <- query_point(
    lat = coords[2],
    lng = coords[1],
    date = date,
    dist = radius_km,
    key = api_key
  )
  
  tibble(point_id = i,
         richness = ifelse(is.null(species), 0, length(species))
         )
  })

richness_df <- bind_rows(results)

# serge with spatial points
points_rich <- points |> 
  left_join(richness_df, by = "point_id")

ggplot() +
  geom_sf(data = klamath, fill = NA, color = "gray50") +
  geom_sf(data = points_rich, aes(color = richness), size = 4) +
  scale_color_viridis(option = "plasma", direction = -1) +
  theme_minimal() +
  labs(title = "Bird Species Richness in the Klamath Basin (API, 10km buffer)",
       color = "Species Richness")
```


Puling 10 years

```{r}
# I have downloaded data from 7 different counties where the klamath basin is located

# Paths
data_dir <- "data-raw/ebird_data/new"
shapefile_path <- "data-raw/ebird_data/klamath_basin_outline.shp"

# Define base filenames (no extension)
counties <- c(
  "ebd_US-CA-015_201501_202507_smp_relJun-2025", # Del Norte
  "ebd_US-CA-023_201501_202507_smp_relJun-2025", # Humboldt
  "ebd_US-CA-049_201501_202507_smp_relJun-2025", # Modoc
  "ebd_US-CA-093_201501_202507_smp_relJun-2025", # Siskiyou
  "ebd_US-CA-105_201501_202507_smp_relJun-2025", # Trinity
  "ebd_US-OR-029_201501_202507_smp_relJun-2025", # Lake
  "ebd_US-OR-037_201501_202507_smp_relJun-2025", # Jackson
  "ebd_US-OR-035_201501_202507_smp_relJun-2025"
)

# Filter all files (skip this section if already done)
# for (base in counties) {
#   ebd_file <- file.path(data_dir, paste0(base, ".txt"))
#   samp_file <- file.path(data_dir, paste0(base, "_sampling.txt"))
# 
#   filt <- auk_ebd(ebd_file, samp_file) |>
#     auk_complete() |>
#     auk_protocol(protocol = c("Traveling", "Stationary")) |>
#     auk_duration(duration = c(0, 120)) |>
#     auk_distance(distance = c(0, 5))
# 
# 
#    auk_filter(
#     filt,
#     file = file.path(data_dir, paste0(base, "_filtered.txt")),
#     sampling_file = file.path(data_dir, paste0(base, "_filtered_sampling.txt")),
#     execute = FALSE,  # Do not run awk from R
#     awk_file = file.path(data_dir, paste0(base, "_filter.awk"))
#   )
# }

# all filtered data
ebd_all <- purrr::map_dfr(
  counties,
  ~ read_ebd(file.path(data_dir, paste0(.x, "_filtered.txt")))
)
```

```{r}
# Read Klamath Basin shapefile
klamath_basin <- st_read("data-raw/klamath_basin_outline/R8_FAC_Klamath_Basin_WFL1.shp") |>
  st_transform(4326)

# Convert EBD to sf points
ebd_all_sf <- ebd_all |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # WGS84

# Ensure both are in the same CRS
klamath_basin <- st_transform(klamath_basin, 4326)

# Keep only points inside the basin
ebd_klamath <- ebd_all_sf[st_within(ebd_all_sf, klamath_basin, sparse = FALSE), ]


ebd_klamath |>
  mutate(year = year(observation_date)) |>
  group_by(year) |>
  summarize(richness = n_distinct(common_name)) |>
  arrange(year)
```

```{r}
# ggplot() +
#   geom_sf(data = klamath_basin, fill = NA, color = "black") +
#   geom_sf(data = ebd_klamath, aes(color = common_name), alpha = 0.5) +
#   theme_minimal()

```


```{r echo=FALSE}

# temporal trend of species richness
ebd_klamath |>  
  mutate(year = year(observation_date)) |> 
  group_by(year) |> 
  summarize(richness = n_distinct(common_name)) |> 
  ggplot(aes(x = year, y = richness)) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "Species Richness in Klamath Basin (2015–2024)",
       x = "Year", y = "Unique Species Observed") +
  theme_minimal()

# All observations
ggplot() +
  geom_sf(data = klamath_basin, fill = NA, color = "black") +
  geom_sf(data = ebd_klamath, alpha = 0.3, color = "blue", size = 0.5) +
  labs(title = "Bird Observations in Klamath Basin") +
  theme_minimal()
```

Sptial Species Richness
```{r echo=FALSE}
# Create hexagonal grid over Klamath Basin
# hex_grid <- st_make_grid(klamath_basin, cellsize = 0.1, square = FALSE)
# hex_grid <- st_sf(geometry = hex_grid)
# 
# # Spatial join
# joined <- st_join(ebd_klamath, hex_grid)
# 
# # Calculate richness per hex
# hex_richness <- joined |> 
#   group_by(geometry) |> 
#   summarize(richness = n_distinct(common_name)) |> 
#   ungroup()
# 
# # Plot
# ggplot(hex_richness) +
#   geom_sf(aes(fill = richness), color = NA) +
#   scale_fill_viridis_c(option = "plasma") +
#   labs(title = "Spatial Species Richness in Klamath Basin",
#        fill = "Richness") +
#   theme_minimal()
```

Top species Observed
```{r}
ebd_klamath |> 
  count(common_name, sort = TRUE) |> 
  slice_max(n, n = 10) |> 
  ggplot(aes(x = reorder(common_name, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Most Observed Species",
       x = "Species", y = "Number of Observations") +
  theme_minimal()
```

Seasonal Bird Species Richness

  - Note that this can be misinterpreted due to the temporal variability of bird watchers

```{r echo=FALSE}
ebd_klamath |> 
  mutate(year = year(observation_date),
         season = case_when(month(observation_date) %in% 3:5 ~ "Spring",
                            month(observation_date) %in% 6:8 ~ "Summer",
                            month(observation_date) %in% 9:11 ~ "Fall",
                            TRUE ~ "Winter")) |> 
  group_by(year, season) |> 
  summarize(richness = n_distinct(common_name), .groups = "drop") |> 
  ggplot(aes(x = year, y = richness, color = season)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  labs(title = "Seasonal Bird Species Richness in the Klamath Basin",
       x = "Year", y = "Species Richness") +
  theme_minimal()

```


```{r echo=FALSE}
# 
ebd_klamath_species_observed <- ebd_klamath |> 
  group_by(sampling_event_identifier) |> 
  mutate(species_observed = n_distinct(common_name))

ggplot(ebd_klamath_species_observed, aes(geometry = geometry, color = species_observed)) +
  geom_sf(alpha = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "Checklist Species Richness in Klamath Basin",
       color = "Species Observed") +
  theme_minimal()

```


```{r echo=FALSE}
ebd_klamath |> 
  mutate(month = month(observation_date, label = TRUE)) |> 
  group_by(month) |> 
  summarize(richness = n_distinct(common_name)) |> 
  ggplot(aes(x = month, y = richness)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Monthly Bird Species Richness",
       x = "Month", y = "Number of Unique Species") +
  theme_minimal()

```

```{r}

```


