---
title: "Species Richness Analysis in the Klamath Basin (2015–2025)"
author: "Badhia Yunes Katz"
date: "July 2025"
output: rmarkdown::github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.width=15, fig.height=10)
library(tidyverse)
library(httr)
library(jsonlite)
library(auk)
library(dplyr)
library(rnaturalearth)
library(sf)
library(ggplot2)
library(readr)
library(viridis)
library(lubridate)
# Sys.setenv(PATH = paste(Sys.getenv("PATH"), "C:/Program Files/Git/usr/bin", sep = ";"))


# api_key <- "ebird_key"
# radius_km <- 80                 
# sample_size <- 15     
# date <- "2024/06/15"  
```

The purpose of this markdown is to analyze and explore bird species richness across the Klamath Basin over the past 10 years using publicly available eBird data.

## Data Source

Data were obtained from the [eBird Basic Dataset (EBD)](https://science.ebird.org/en/use-ebird-data/download-ebird-data-products) via custom downloads. The following counties intersecting the Klamath Basin were included:

  - California: Del Norte (CA-015), Humboldt (CA-023), Modoc (CA-049), Siskiyou (CA-093), Trinity (CA-105)

  - Oregon: Jackson (OR-029), Lake (OR-037), Klamath (OR-035)

Each download included:

  - Observation data (.txt)

  - Sampling event metadata (_sampling.txt)

  - Data filtered using the [auk](https://cornelllabofornithology.github.io/auk/) R package


## Spatial Filtering

A shapefile representing the Klamath Basin watershed boundary was used to:

  - Limit observations to only those falling within the basin

  - Ensure accurate spatial representation of species richness and effort

Spatial filtering was performed using the sf R package, with observations reprojected and subset using spatial joins.

## Limitations and Biases

While eBird provides a vast and valuable source of bird observation data, several **limitations and biases** must be considered when interpreting species richness in the Klamath Basin:

### Spatial and Temporal Bias

- **Uneven geographic coverage**: Observations are concentrated near roads, towns, protected areas, or known birding hotspots rather than being evenly distributed across the basin.
- **Accessibility**: Remote, rugged, or private lands are under-surveyed due to difficulty of access.
- **Temporal bias**: Observations are skewed toward **weekends**, holidays, and the **month of May**, when birding activity typically increases due to spring migration and favorable weather.

### Observer Skill and Experience

- Bird detection and identification are highly dependent on the skill level of the observer:
  - **Novice observers** may miss or misidentify species, especially cryptic or similar-looking birds.
  - **Experienced birders** may detect more species per checklist, leading to inflated richness in certain areas.


### Personal Preference and Targeting

- Observer decisions introduce subjectivity:
  - Birders may intentionally seek out **rare** or **charismatic** species.
  - Areas near homes, parks, or favorite hotspots may be surveyed more often.
  - “Chasing” behavior (targeting recent rare sightings) may artificially inflate species counts at specific locations.


### Checklist Effort and Metadata Variability

- Despite quality filters (complete checklists, effort constraints), variability remains in:
  - **Number of observers**
  - **Start time of surveys**
  - **Use of playback, feeders, or attractants**

These factors influence the likelihood of species detection.

### Presence-Only Limitations

- eBird is a **presence-only** dataset.
  - A species not reported on a checklist does **not imply absence**—it may simply have gone undetected.
  - Species with low detectability or limited seasonal presence may be underrepresented.



```{r eval=FALSE, include=FALSE}
# using ebird API

# Load Klamath Basin polygon
klamath_basin <- st_read("data-raw/klamath_basin_outline/R8_FAC_Klamath_Basin_WFL1.shp")
klamath <- st_transform(klamath_basin, crs = 4326)

# picking sample points  
points <- st_sample(klamath, size = sample_size, type = "random") |> 
  st_sf() |> 
  mutate(point_id = row_number())

# query eBird points
query_point <- function(lat, lng, date, dist, key) {
  url <- paste0(
    "https://api.ebird.org/v2/data/obs/geo/historic/",
    date,
    "?lat=", lat,
    "&lng=", lng,
    "&dist=", dist,
    "&maxResults=500")
  
  resp <- GET(url, add_headers("X-eBirdApiToken" = key))
  if (http_status(resp)$category != "Success") return(NULL)
  
  dat <- fromJSON(content(resp, "text", encoding = "UTF-8"))
  if (length(dat) == 0) return(NULL)
  
  unique_species <- unique(dat$comName)
  return(unique_species)
}

# richness 
results <- lapply(seq_len(nrow(points)), function(i) {
  coords <- st_coordinates(points[i, ])
  species <- query_point(
    lat = coords[2],
    lng = coords[1],
    date = date,
    dist = radius_km,
    key = api_key
  )
  
  tibble(point_id = i,
         richness = ifelse(is.null(species), 0, length(species))
         )
  })

richness_df <- bind_rows(results)

# serge with spatial points
points_rich <- points |> 
  left_join(richness_df, by = "point_id")

ggplot() +
  geom_sf(data = klamath, fill = NA, color = "gray50") +
  geom_sf(data = points_rich, aes(color = richness), size = 4) +
  scale_color_viridis(option = "plasma", direction = -1) +
  theme_minimal() +
  labs(title = "Bird Species Richness in the Klamath Basin (API, 10km buffer)",
       color = "Species Richness")
```


Puling 10 years

```{r include=FALSE}
# I have downloaded data from 7 different counties where the klamath basin is located

# Paths
data_dir <- "data-raw/ebird_data/new"
shapefile_path <- "data-raw/ebird_data/klamath_basin_outline.shp"

# Define base filenames (no extension)
# counties <- c(
#   "ebd_US-CA-015_201501_202507_smp_relJun-2025", # Del Norte
#   "ebd_US-CA-023_201501_202507_smp_relJun-2025", # Humboldt
#   "ebd_US-CA-049_201501_202507_smp_relJun-2025", # Modoc
#   "ebd_US-CA-093_201501_202507_smp_relJun-2025", # Siskiyou
#   "ebd_US-CA-105_201501_202507_smp_relJun-2025", # Trinity
#   "ebd_US-OR-029_201501_202507_smp_relJun-2025", # Lake
#   "ebd_US-OR-037_201501_202507_smp_relJun-2025", # Jackson
#   "ebd_US-OR-035_201501_202507_smp_relJun-2025"
# )

# Filter all files (skip this section if already done)
# for (base in counties) {
#   ebd_file <- file.path(data_dir, paste0(base, ".txt"))
#   samp_file <- file.path(data_dir, paste0(base, "_sampling.txt"))
# 
#   filt <- auk_ebd(ebd_file, samp_file) |>
#     auk_complete() |>
#     auk_protocol(protocol = c("Traveling", "Stationary")) |>
#     auk_duration(duration = c(0, 120)) |>
#     auk_distance(distance = c(0, 5))
# 
# 
#    auk_filter(
#     filt,
#     file = file.path(data_dir, paste0(base, "_filtered.txt")),
#     sampling_file = file.path(data_dir, paste0(base, "_filtered_sampling.txt")),
#     execute = FALSE,  # Do not run awk from R
#     awk_file = file.path(data_dir, paste0(base, "_filter.awk"))
#   )
# }

# all filtered data
# ebd_all <- purrr::map_dfr(
#   counties,
#   ~ read_ebd(file.path(data_dir, paste0(.x, "_filtered.txt")))
# )

# Save combined data to disk so that map_dfr does not have to be ran every time
# saveRDS(ebd_all, file = "data/ebd_klamath_combined.rds")

# Load combined data
ebd_all <- readRDS("data/ebd_klamath_combined.rds")

```

```{r include=FALSE}
# Read Klamath Basin shapefile
klamath_basin <- st_read("data-raw/klamath_basin_outline/R8_FAC_Klamath_Basin_WFL1.shp") |>
  st_transform(4326)

# Convert EBD to sf points
ebd_all_sf <- ebd_all |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # WGS84

# Ensure both are in the same CRS
klamath_basin <- st_transform(klamath_basin, 4326)

# Keep only points inside the basin
ebd_klamath <- ebd_all_sf[st_within(ebd_all_sf, klamath_basin, sparse = FALSE), ]


ebd_klamath |>
  mutate(year = year(observation_date)) |>
  group_by(year) |>
  summarize(richness = n_distinct(common_name)) |>
  arrange(year)
```

```{r}
# ggplot() +
#   geom_sf(data = klamath_basin, fill = NA, color = "black") +
#   geom_sf(data = ebd_klamath, aes(color = common_name), alpha = 0.5) +
#   theme_minimal()

```

## Temporal Trends in Species Richness

This plot shows the total number of unique species recorded per year. Peaks may correspond to years with higher survey effort, observer participation, or migration events.

```{r echo=FALSE}
# temporal trend of species richness
ebd_klamath |>  
  mutate(year = year(observation_date)) |> 
  group_by(year) |> 
  summarize(richness = n_distinct(common_name)) |> 
  ggplot(aes(x = year, y = richness)) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "Species Richness in Klamath Basin (2015–2024)",
       x = "Year", y = "Unique Species Observed") +
  theme_minimal()
```

## Spatial coverage map

This map displays all checklists within the Klamath Basin. Denser clusters likely reflect accessibility (e.g., roads, hotspots) and observer preferences.

```{r echo=FALSE}
# All observations
ggplot() +
  geom_sf(data = klamath_basin, fill = NA, color = "black") +
  geom_sf(data = ebd_klamath, alpha = 0.3, color = "blue", size = 0.5) +
  labs(title = "Bird Observations in Klamath Basin") +
  theme_minimal()
```

```{r echo=FALSE}
# Create hexagonal grid over Klamath Basin
# hex_grid <- st_make_grid(klamath_basin, cellsize = 0.1, square = FALSE)
# hex_grid <- st_sf(geometry = hex_grid)
# 
# # Spatial join
# joined <- st_join(ebd_klamath, hex_grid)
# 
# # Calculate richness per hex
# hex_richness <- joined |> 
#   group_by(geometry) |> 
#   summarize(richness = n_distinct(common_name)) |> 
#   ungroup()
# 
# # Plot
# ggplot(hex_richness) +
#   geom_sf(aes(fill = richness), color = NA) +
#   scale_fill_viridis_c(option = "plasma") +
#   labs(title = "Spatial Species Richness in Klamath Basin",
#        fill = "Richness") +
#   theme_minimal()
```

## Top 10 Most Frequently Observed Species

These species are the most commonly reported across all checklists, often due to being common, conspicuous, or resident year-round.

The most common species is American Robin

```{r}
ebd_klamath |> 
  count(common_name, sort = TRUE) |> 
  slice_max(n, n = 10) |> 
  ggplot(aes(x = reorder(common_name, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Most Observed Species",
       x = "Species", y = "Number of Observations") +
  theme_minimal()
```

Seasonal Bird Species Richness

  - Note that this can be misinterpreted due to the temporal variability of bird watchers

```{r echo=FALSE}
ebd_klamath |> 
  mutate(year = year(observation_date),
         season = case_when(month(observation_date) %in% 3:5 ~ "Spring",
                            month(observation_date) %in% 6:8 ~ "Summer",
                            month(observation_date) %in% 9:11 ~ "Fall",
                            TRUE ~ "Winter")) |> 
  group_by(year, season) |> 
  summarize(richness = n_distinct(common_name), .groups = "drop") |> 
  ggplot(aes(x = year, y = richness, color = season)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  labs(title = "Seasonal Bird Species Richness in the Klamath Basin",
       x = "Year", y = "Species Richness") +
  theme_minimal()

```


```{r echo=FALSE}
# 
ebd_klamath_species_observed <- ebd_klamath |> 
  group_by(sampling_event_identifier) |> 
  mutate(species_observed = n_distinct(common_name))

ggplot(ebd_klamath_species_observed, aes(geometry = geometry, color = species_observed)) +
  geom_sf(alpha = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "Checklist Species Richness in Klamath Basin",
       color = "Species Observed") +
  theme_minimal()

```

## Monthly Richness Trends

Richness peaks in spring and early summer, again corresponding to migration and peak observation periods

```{r echo=FALSE}
ebd_klamath |> 
  mutate(month = month(observation_date, label = TRUE)) |> 
  group_by(month) |> 
  summarize(richness = n_distinct(common_name)) |> 
  ggplot(aes(x = month, y = richness)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Monthly Bird Species Richness",
       x = "Month", y = "Number of Unique Species") +
  theme_minimal()

```

```{r}

```


