---
title: "eBird Data Exploration"
author: "Badhia Yunes Katz"
date: "July 2025"
output: rmarkdown::github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.width=15, fig.height=10)
library(tidyverse)
library(httr)
library(jsonlite)
library(auk)
```

The purpose of this markdown is to bird data. This data was sourced from [CornellLab eBird Data and Tools Site](https://science.ebird.org/en/use-ebird-data). 

Data product citation:
Fink, D., T. Auer, A. Johnston, M. Strimas-Mackey, S. Ligocki, O. Robinson, W. Hochachka, L. Jaromczyk, C. Crowley, K. Dunham, A. Stillman, I. Davies, A. Rodewald, V. Ruiz-Gutierrez, C. Wood. 2023. eBird Status and Trends, Data Version: 2022; Released: 2023. Cornell Lab of Ornithology, Ithaca, New York. https://doi.org/10.2173/ebirdst.2022

### Overview


```{r}
bbox_url <- "https://api.ebird.org/v2/data/obs/geo/recent"

api_key = Sys.getenv("ebird_key")

response <- GET(
  url = bbox_url,
  query = list(
    lat = 42.0,
    lng = -121.5,
    dist = 50,
    maxResults = 500
  ),
  add_headers("X-eBirdApiToken" = api_key)
)

# Parse results
data <- fromJSON(content(response, "text", encoding = "UTF-8"))

# View species
unique_species <- unique(data$comName)
print(unique_species)
```

