---
title: "Klamath Bird Data - Synthesized"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(auk)
library(dplyr)
library(rnaturalearth)
library(sf)
library(ggplot2)
library(readr)
library(viridis)
library(lubridate)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12, 
  fig.height = 12
)

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14)
    )
)
```

## Questions

1.  Do we want individual species tracked to guild?
2.  Should we add a geographic component to the dataset?
3.  Can we provide more specificity in guild structure?

## Build Dataset {.tabset}

We compiled eBird observations from 2015–2025 for the Klamath Basin watershed. These data were merged with BIRDBASE, which provides information on each species’ primary habitat and diet. Additional information on waterbirds was drawn from Donnelly et al. (2022). Further details on each dataset are provided in the following tabs.

Below, we explore potential groupings of bird species based on these combined sources [in the Bird Groupings section](#bird-groupings)

### eBird

##### Data Source

Data were obtained from the [eBird Basic Dataset (EBD)](https://science.ebird.org/en/use-ebird-data/download-ebird-data-products) via custom downloads. The following counties intersecting the Klamath Basin were included:

-   California: Del Norte (CA-015), Humboldt (CA-023), Modoc (CA-049), Siskiyou (CA-093), Trinity (CA-105)

-   Oregon: Jackson (OR-029), Lake (OR-037), Klamath (OR-035)

Each download included:

-   Observation data (.txt)

-   Sampling event metadata (\_sampling.txt)

-   Data filtered using the [auk](https://cornelllabofornithology.github.io/auk/) R package

##### Spatial Filtering

A shapefile representing the Klamath Basin watershed boundary was used to:

-   Limit observations to only those falling within the basin

-   Ensure accurate spatial representation of species richness and effort

Spatial filtering was performed using the sf R package, with observations reprojected and subset using spatial joins.

#### Limitations and Biases

While eBird provides a vast and valuable source of bird observation data, several **limitations and biases** must be considered when interpreting species richness in the Klamath Basin:

#### Spatial and Temporal Bias

-   **Uneven geographic coverage**: Observations are concentrated near roads, towns, protected areas, or known birding hotspots rather than being evenly distributed across the basin.
-   **Accessibility**: Remote, rugged, or private lands are under-surveyed due to difficulty of access.
-   **Temporal bias**: Observations are skewed toward **weekends**, holidays, and the **month of May**, when birding activity typically increases due to spring migration and favorable weather.

#### Observer Skill and Experience

-   Bird detection and identification are highly dependent on the skill level of the observer:
-   **Novice observers** may miss or misidentify species, especially cryptic or similar-looking birds.
-   **Experienced birders** may detect more species per checklist, leading to inflated richness in certain areas.

#### Personal Preference and Targeting

-   Observer decisions introduce subjectivity:
-   Birders may intentionally seek out **rare** or **charismatic** species.
-   Areas near homes, parks, or favorite hotspots may be surveyed more often.
-   “Chasing” behavior (targeting recent rare sightings) may artificially inflate species counts at specific locations.

##### Checklist Effort and Metadata Variability

-   Despite quality filters (complete checklists, effort constraints), variability remains in:
-   **Number of observers**
-   **Start time of surveys**
-   **Use of playback, feeders, or attractants**

These factors influence the likelihood of species detection.

#### Presence-Only Limitations

-   eBird is a **presence-only** dataset.
-   A species not reported on a checklist does **not imply absence**—it may simply have gone undetected.
-   Species with low detectability or limited seasonal presence may be underrepresented.

[This markdown follows ebird best practices](https://ebird.github.io/ebird-best-practices/)

```{r eval=FALSE, include=FALSE}
# using ebird API

# Load Klamath Basin polygon
klamath_basin <- st_read("data-raw/klamath_basin_outline/R8_FAC_Klamath_Basin_WFL1.shp")
klamath <- st_transform(klamath_basin, crs = 4326)

# picking sample points  
points <- st_sample(klamath, size = sample_size, type = "random") |> 
  st_sf() |> 
  mutate(point_id = row_number())

# query eBird points
query_point <- function(lat, lng, date, dist, key) {
  url <- paste0(
    "https://api.ebird.org/v2/data/obs/geo/historic/",
    date,
    "?lat=", lat,
    "&lng=", lng,
    "&dist=", dist,
    "&maxResults=500")
  
  resp <- GET(url, add_headers("X-eBirdApiToken" = key))
  if (http_status(resp)$category != "Success") return(NULL)
  
  dat <- fromJSON(content(resp, "text", encoding = "UTF-8"))
  if (length(dat) == 0) return(NULL)
  
  unique_species <- unique(dat$comName)
  return(unique_species)
}

# richness 
results <- lapply(seq_len(nrow(points)), function(i) {
  coords <- st_coordinates(points[i, ])
  species <- query_point(
    lat = coords[2],
    lng = coords[1],
    date = date,
    dist = radius_km,
    key = api_key
  )
  
  tibble(point_id = i,
         richness = ifelse(is.null(species), 0, length(species))
  )
})

richness_df <- bind_rows(results)

# serge with spatial points
points_rich <- points |> 
  left_join(richness_df, by = "point_id")

ggplot() +
  geom_sf(data = klamath, fill = NA, color = "gray50") +
  geom_sf(data = points_rich, aes(color = richness), size = 4) +
  scale_color_viridis(option = "plasma", direction = -1) +
  theme_minimal() +
  labs(title = "Bird Species Richness in the Klamath Basin (API, 10km buffer)",
       color = "Species Richness")
```

```{r include=FALSE}
# I have downloaded data from 7 different counties where the klamath basin is located

# Paths
data_dir <- "data-raw/ebird_data/new"
shapefile_path <- "data-raw/ebird_data/klamath_basin_outline.shp"

# Define base filenames (no extension)
# counties <- c(
#   "ebd_US-CA-015_201501_202507_smp_relJun-2025", # Del Norte
#   "ebd_US-CA-023_201501_202507_smp_relJun-2025", # Humboldt
#   "ebd_US-CA-049_201501_202507_smp_relJun-2025", # Modoc
#   "ebd_US-CA-093_201501_202507_smp_relJun-2025", # Siskiyou
#   "ebd_US-CA-105_201501_202507_smp_relJun-2025", # Trinity
#   "ebd_US-OR-029_201501_202507_smp_relJun-2025", # Lake
#   "ebd_US-OR-037_201501_202507_smp_relJun-2025", # Jackson
#   "ebd_US-OR-035_201501_202507_smp_relJun-2025"
# )

# Filter all files (skip this section if already done)
# for (base in counties) {
#   ebd_file <- file.path(data_dir, paste0(base, ".txt"))
#   samp_file <- file.path(data_dir, paste0(base, "_sampling.txt"))
# 
#   filt <- auk_ebd(ebd_file, samp_file) |>
#     auk_complete() |>
#     auk_protocol(protocol = c("Traveling", "Stationary")) |>
#     auk_duration(duration = c(0, 120)) |>
#     auk_distance(distance = c(0, 5))
# 
# 
#    auk_filter(
#     filt,
#     file = file.path(data_dir, paste0(base, "_filtered.txt")),
#     sampling_file = file.path(data_dir, paste0(base, "_filtered_sampling.txt")),
#     execute = FALSE,  # Do not run awk from R
#     awk_file = file.path(data_dir, paste0(base, "_filter.awk"))
#   )
# }

# all filtered data
# ebd_all <- purrr::map_dfr(
#   counties,
#   ~ read_ebd(file.path(data_dir, paste0(.x, "_filtered.txt")))
# )

# Save combined data to disk so that map_dfr does not have to be ran every time
# saveRDS(ebd_all, file = "data/ebd_klamath_combined.rds")

# Load combined data
ebd_all <- readRDS(here::here("data", "ebd_klamath_combined.rds"))

```

```{r include=FALSE}
# Read Klamath Basin shapefile
klamath_basin <- st_read(here::here("data-raw", "klamath_basin_outline/R8_FAC_Klamath_Basin_WFL1.shp")) |>
  st_transform(4326)

# Convert EBD to sf points
ebd_all_sf <- ebd_all |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # WGS84

# Ensure both are in the same CRS
klamath_basin <- st_transform(klamath_basin, 4326)

# Keep only points inside the basin
ebd_klamath <- ebd_all_sf[st_within(ebd_all_sf, klamath_basin, sparse = FALSE), ]


ebd_klamath |>
  mutate(year = year(observation_date)) |>
  group_by(year) |>
  summarize(richness = n_distinct(common_name)) |>
  arrange(year)
```

#### Temporal Trends in Species Richness

This plot shows the total number of unique species recorded per year. Peaks may correspond to years with higher survey effort, observer participation, or migration events.

```{r echo=FALSE}
# temporal trend of species richness
ebd_klamath |>  
  mutate(year = year(observation_date)) |> 
  group_by(year) |> 
  summarize(richness = n_distinct(common_name)) |> 
  ggplot(aes(x = year, y = richness)) +
  geom_line(size = 1.2) +
  geom_point() +
  scale_x_continuous(
    breaks = seq(2015, 2024, by = 1),
    labels = seq(2015, 2024, by = 1)) +
  labs(title = "Species Richness in Klamath Basin (2015–2024)",
       x = "Year", y = "Unique Species Observed") +
  theme_minimal()
```

#### Spatial coverage map

This map displays all checklists within the Klamath Basin. Denser clusters likely reflect accessibility (e.g., roads, hotspots) and observer preferences.

```{r echo=FALSE}
# All observations
ggplot() +
  geom_sf(data = klamath_basin, fill = NA, color = "black") +
  geom_sf(data = ebd_klamath, alpha = 0.3, color = "blue", size = 0.5) +
  labs(title = "Bird Observations in Klamath Basin") +
  theme_minimal()
```

```{r echo=FALSE}
# Create hexagonal grid over Klamath Basin
# hex_grid <- st_make_grid(klamath_basin, cellsize = 0.1, square = FALSE)
# hex_grid <- st_sf(geometry = hex_grid)
# 
# # Spatial join
# joined <- st_join(ebd_klamath, hex_grid)
# 
# # Calculate richness per hex
# hex_richness <- joined |> 
#   group_by(geometry) |> 
#   summarize(richness = n_distinct(common_name)) |> 
#   ungroup()
# 
# # Plot
# ggplot(hex_richness) +
#   geom_sf(aes(fill = richness), color = NA) +
#   scale_fill_viridis_c(option = "plasma") +
#   labs(title = "Spatial Species Richness in Klamath Basin",
#        fill = "Richness") +
#   theme_minimal()
```

#### Top 10 Most Frequently Observed Species

These species are the most commonly reported across all checklists, often due to being common, conspicuous, or resident year-round.

The most common species is American Robin

```{r echo=FALSE}
ebd_klamath |> 
  count(common_name, sort = TRUE) |> 
  slice_max(n, n = 10) |> 
  ggplot(aes(x = reorder(common_name, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Most Observed Species",
       x = "Species", y = "Number of Observations") +
  theme_minimal()
```

Seasonal Bird Species Richness

Spring is the season with more observations from bird watchers, and winter is the lowest. This is not surprising due to the frequency of when people spend time outside during these seasons.

-   Note that this can be misinterpreted due to the temporal variability of bird watchers

```{r echo=FALSE}
ebd_klamath |> 
  mutate(year = year(observation_date),
         season = case_when(month(observation_date) %in% 3:5 ~ "Spring",
                            month(observation_date) %in% 6:8 ~ "Summer",
                            month(observation_date) %in% 9:11 ~ "Fall",
                            TRUE ~ "Winter")) |> 
  group_by(year, season) |> 
  summarize(richness = n_distinct(common_name), .groups = "drop") |> 
  ggplot(aes(x = year, y = richness, color = season)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = seq(2015, 2024, by = 1),
    labels = seq(2015, 2024, by = 1)) +
  labs(title = "Seasonal Bird Species Richness in the Klamath Basin",
       x = "Year", y = "Species Richness") +
  theme_minimal()

```

Checklist Species Richness in Klamath Basin

species_observed value in your code represents the number of unique species recorded in each individual eBird checklist

```{r echo=FALSE}
# TODO play around with color scale
ebd_klamath_species_observed <- ebd_klamath |> 
  group_by(sampling_event_identifier) |> 
  mutate(species_observed = n_distinct(common_name))

#How many species are typically seen per checklist
ggplot(ebd_klamath_species_observed, aes(geometry = geometry, color = species_observed)) +
  geom_sf(alpha = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(
    title = "Checklist Species Richness in the Klamath Basin",
    subtitle = "Each point represents a complete checklist (2015–2024)") +
  theme_minimal(base_size = 12)

```

#### Monthly Richness Trends

Richness peaks in spring and early summer, again corresponding to migration and peak observation periods

```{r echo=FALSE}
ebd_klamath |> 
  mutate(month = month(observation_date, label = TRUE)) |> 
  group_by(month) |> 
  summarize(richness = n_distinct(common_name)) |> 
  ggplot(aes(x = month, y = richness)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Monthly Bird Species Richness",
       x = "Month", y = "Number of Unique Species") +
  theme_minimal()

```

```{r}
ebd_all <- readRDS(here::here("data", "ebd_klamath_combined.rds")) |> glimpse()
```

### BIRDBASE

BIRDBASE [^1] provides a thorough, literature based review of all bird species and includes habitat and foraging information that can be used to assign birds to guilds.

[^1]: Şekercioğlu, Çağan H; Kittelberger, Kyle D.; Mota, Flávia; Buxton, Amy N.; Orton, Nikolas; DeNiro, Adara; et al. (2025). BIRDBASE: A Global Database of Avian Biogeography, Conservation, Ecology and Life History Traits. figshare. Dataset. <https://doi.org/10.6084/m9.figshare.27051040.v1>

The `primary habitat` and `primary diet` values were used but it's worth noting that BIRDBASE provides information on how each bird species fits into each type. This could help if a more involved analysis is requested.

```{r}
birdbase <- readxl::read_excel(here::here('data-raw', 'BIRDBASE v2025.1 Sekercioglu et al. Final.xlsx'), sheet = "Data", skip = 1) |> 
  janitor::clean_names() |> glimpse()
```

3.  Merge ebird with BIRDBASE to get guild information:

The two datasets can easily be merged by the `scientific name`. The final dataset was heavily truncated to include on the data that is needed for this analysis.

```{r}
ebd_trunc <- ebd_all |> 
  select(common_name, scientific_name, observation_count, latitude, longitude, observation_date)

birdbase_trunc <- birdbase |> 
  select(scientific_name = e_bird_clements_v2024b, primary_diet, primary_habitat, species, genus, order)

ebd_birdbase <- ebd_trunc |> 
  left_join(birdbase_trunc) 
```

### Waterbird Niche

Based off of table 2 of Donnelly et al., 2022[^2]. This provides additional information for waterbirds.

[^2]: Donnelly, J. P., Moore, J. N., Casazza, M. L., & Coons, S. P. (2022). Functional wetland loss drives emerging risks to waterbird migration networks. *Frontiers in Ecology and Evolution, 10*, 844278. <https://doi.org/10.3389/fevo.2022.844278>

```{r}
waterbirds <- tribble(
  ~waterbird_niche,              ~common_name,
  # Shorebirds
  "Shorebirds",        "American Avocet",
  "Shorebirds",        "Black-necked Stilt",
  "Shorebirds",        "Dunlin",
  "Shorebirds",        "Greater Yellowlegs",
  "Shorebirds",        "Lesser Yellowlegs",
  "Shorebirds",        "Long-billed Dowitcher",
  "Shorebirds",        "Marbled Godwit",
  "Shorebirds",        "Western Sandpiper",
  "Shorebirds",        "Willet",
  "Shorebirds",        "Wilson's Phalarope",
  "Shorebirds",        "Wilson's Snipe",
  "Shorebirds",        "Whimbrel",
  
  # Dabbling ducks
  "Dabbling ducks",    "American Wigeon",
  "Dabbling ducks",    "Cinnamon Teal",
  "Dabbling ducks",    "Gadwall",
  "Dabbling ducks",    "Green-winged Teal",
  "Dabbling ducks",    "Mallard",
  "Dabbling ducks",    "Northern Pintail",
  "Dabbling ducks",    "Northern Shoveler",
  
  # Diving ducks
  "Diving ducks",      "Barrow's Goldeneye", 
  "Diving ducks",      "Common Goldeneye",
  "Diving ducks",      "Bufflehead",
  "Diving ducks",      "Canvasback",
  "Diving ducks",      "Greater Scaup", 
  "Diving ducks",      "Lesser Scaup",
  "Diving ducks",      "Redhead",
  "Diving ducks",      "Ring-necked Duck",
  "Diving ducks",      "Ruddy Duck",
  
  # Other waterbirds
  "Other waterbirds",  "American Bittern",
  "Other waterbirds",  "American Coot",
  "Other waterbirds",  "Black Tern",
  "Other waterbirds",  "Eared Grebe",
  "Other waterbirds",  "Least Bittern",
  "Other waterbirds",  "Western Grebe",
  "Other waterbirds",  "White-faced Ibis"
) 

knitr::kable(waterbirds)

```

```{r}
ebd_birdbase_with_waterbirds <- ebd_birdbase |> 
  left_join(waterbirds) |> glimpse()
```

### RTR/USGS Piscivorous waterbirds counts

Herons =  Great Blue Herons, Great Egrets, and Black-crowned Night
Gulls = California and Ring-billed gulls 

**Sources:**

-   2009 - 2020: Table S.1 of Evans et al., 2022[^3]
-   2021-2023: Table 1 of Banet et al., 2024[^4]
-   2024: Table 1 of Banet et al., 2025[^5]

[^3]: Evans, A. F., Payton, Q., Banet, N., Cramer, B. M., Kelsey, C., & Hewitt, D. A. (2022). Avian predation on juvenile and adult Lost River and Shortnose suckers: An updated multi-predator species evaluation. North American Journal of Fisheries Management. <https://doi.org/10.1002/nafm.10838>; <https://birdresearchnw.org/wp-content/uploads/2025/11/Evans20et20al20202220Sucker20Predation.pdf>

[^4]: Banet, N., Payton, Q., Evans, A., Krause, J., Hayes, B., Paul-Wilson, R., & Benham, E. (2024). Predation of Lost River and Shortnose suckers by piscivorous colonial waterbirds in the Upper Klamath Basin: An analysis of predation effects in 2021-2023. U.S. Bureau of Reclamation https://birdresearchnw.org/wp-content/uploads/2025/11/Avian20Predation20on20UKB20Suckers_Summary20Report202021-2023.pdf

[^5]: Banet, N., Payton, Q., Devincenzi, D., Evans, A., Krause, J., Hayes, B., Paul-Wilson, R., & Benham, E. (2025). Predation of Lost River and Shortnose suckers by piscivorous colonial waterbirds in the Upper Klamath Basin: An analysis of predation effects in 2024 (Final report). U.S. Bureau of Reclamation https://birdresearchnw.org/wp-content/uploads/2025/11/Avian-Predation-Final-Summary-Report_2024.pdf



```{r}
piscivorous_waterbirds  <- tribble(
  ~nesting_location,     ~common_name,          ~year,    ~count, ~source,
  
  # UKL
  "Upper Klamath Lake", "American White Pelican",  2009, 438, "Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant", 2009, 1538,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican",  2010, 81,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant", 2010, 293,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican", 2011 , 14,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant",2011 ,373 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican",2012  ,255 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant",2012 , 885,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican", 2013 ,152 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant",2013 ,1071 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican", 2014 ,247 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant",2014 ,1133 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican",  2015,492 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant",2015 ,515 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican", 2018 ,510 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant",2018 ,544 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican",  2019, 129,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant", 2019, 710,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican", 2020 ,391 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "Double-crested Cormorant",2020 ,416 ,"Table S.1 of Evans et al., 2022",
  "Upper Klamath Lake", "American White Pelican", 2022 , 97,"Table 1 of Banet et al., 2024",
  "Upper Klamath Lake", "Double-crested Cormorant", 2022, 158,"Table 1 of Banet et al., 2024",
  "Upper Klamath Lake", "Herons", 2022, NA,"Table 1 of Banet et al., 2024", # colony was active but not counted
  "Upper Klamath Lake", "American White Pelican", 2023 , 232,"Table 1 of Banet et al., 2024",
  "Upper Klamath Lake", "Double-crested Cormorant", 2023, 370,"Table 1 of Banet et al., 2024",
  "Upper Klamath Lake", "Herons", 2023, 574,"Table 1 of Banet et al., 2024",
  "Upper Klamath Lake", "American White Pelican", 2024 , 312,"Table 1 of Banet et al., 2025",
  "Upper Klamath Lake", "Double-crested Cormorant", 2024, 301,"Table 1 of Banet et al., 2025",
  "Upper Klamath Lake", "Herons", 2024, 450,"Table 1 of Banet et al., 2025",

  
  # Clear Lake  
  "Clear Lake Reservoir", "American White Pelican", 2009, 2325,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant", 2009, 172,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican",2010 ,722 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant", 2010,0 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican", 2011,128 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant",2011 ,77 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican",2012 ,510 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant",2012 ,286 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican",2013 ,1175 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant",2013 ,82 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican", 2014,296 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant",2014 ,197 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican", 2015, 503,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant",2015 ,71 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican",2018 ,631 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant",2018 ,162 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican", 2019,58 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant",2019 ,136 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican", 2020,1166 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "Double-crested Cormorant",2020 ,200 ,"Table S.1 of Evans et al., 2022",
  "Clear Lake Reservoir", "American White Pelican", 2021 ,1094 ,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Double-crested Cormorant", 2021,110 ,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Caspian Tern", 2021, 152,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Herons", 2021, NA ,"Table 1 of Banet et al., 2024", # colony was active but not counted
  "Clear Lake Reservoir", "Gulls", 2021, NA ,"Table 1 of Banet et al., 2024", # colony was active but not counted
  "Clear Lake Reservoir", "American White Pelican", 2022 , 509,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Double-crested Cormorant", 2022,114 ,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Caspian Tern", 2022, 0,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Herons", 2022, NA,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Gulls", 2022, NA,"Table 1 of Banet et al., 2024", # colony was active but not counted
  "Clear Lake Reservoir", "American White Pelican", 2023 , 1621,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Double-crested Cormorant", 2023,163 ,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Caspian Tern", 2023, 202,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Herons", 2023, 15,"Table 1 of Banet et al., 2024",
  "Clear Lake Reservoir", "Gulls", 2023, NA,"Table 1 of Banet et al., 2024", # colony was active but not counted
  "Clear Lake Reservoir", "American White Pelican", 2024 ,1422 ,"Table 1 of Banet et al., 2025",
  "Clear Lake Reservoir", "Double-crested Cormorant", 2024,182 ,"Table 1 of Banet et al., 2025",
  "Clear Lake Reservoir", "Herons", 2024,29 ,"Table 1 of Banet et al., 2025",
  "Clear Lake Reservoir", "Caspian Tern", 2024,610 ,"Table 1 of Banet et al., 2025",
  "Clear Lake Reservoir", "Gulls", 2024, 3872,"Table 1 of Banet et al., 2025",
  
  # Sheepy Lake
  "Sheepy Lake", "American White Pelican",2018 , 199,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "Double-crested Cormorant",2018 ,435 ,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "Caspian Tern",2018 , 319,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "American White Pelican", 2019,34 ,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "Double-crested Cormorant",2019 ,412 ,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "Caspian Tern",2019 , 270,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "American White Pelican", 2020,415 ,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "Double-crested Cormorant",2020 , 366,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "Caspian Tern",2020 , 544,"Table S.1 of Evans et al., 2022",
  "Sheepy Lake", "American White Pelican", 2021 ,169 ,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Double-crested Cormorant", 2021,420 ,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Caspian Tern", 2021, 324,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Gulls", 2021,NA ,"Table 1 of Banet et al., 2024", # colony was active but not counted
  "Sheepy Lake", "American White Pelican", 2022 ,40 ,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Double-crested Cormorant", 2022,723 ,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Caspian Tern", 2022, 277,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Gulls", 2022,NA ,"Table 1 of Banet et al., 2024",# colony was active but not counted
  "Sheepy Lake", "American White Pelican", 2023 ,239 ,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Double-crested Cormorant", 2023,680 ,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Caspian Tern", 2023, 290,"Table 1 of Banet et al., 2024",
  "Sheepy Lake", "Gulls", 2023,NA ,"Table 1 of Banet et al., 2024",# colony was active but not counted
  "Sheepy Lake", "American White Pelican", 2024 ,302 ,"Table 1 of Banet et al., 2025",
  "Sheepy Lake", "Double-crested Cormorant", 2024,787 ,"Table 1 of Banet et al., 2025",
  "Sheepy Lake", "Caspian Tern", 2024, 154,"Table 1 of Banet et al., 2025",
  "Sheepy Lake", "Gulls", 2024,3611 ,"Table 1 of Banet et al., 2025",
  
  # Klamath River
   "Klamath River", "Herons", 2024, 140,"Table 1 of Banet et al., 2025",

  # Tule Lake
  "Tule Lake", "American White Pelican",2018 ,36 ,"Table S.1 of Evans et al., 2022",
  "Tule Lake", "Double-crested Cormorant",2018 ,517 ,"Table S.1 of Evans et al., 2022",
  "Tule Lake", "Caspian Tern",2018 , 775,"Table S.1 of Evans et al., 2022",
  "Tule Lake", "American White Pelican", 2019,15 ,"Table S.1 of Evans et al., 2022",
  "Tule Lake", "Double-crested Cormorant",2019 ,263 ,"Table S.1 of Evans et al., 2022",
  "Tule Lake", "Caspian Tern",2019 , 228,"Table S.1 of Evans et al., 2022",
  "Tule Lake", "American White Pelican", 2020,15 ,"Table S.1 of Evans et al., 2022",
  "Tule Lake", "Double-crested Cormorant",2020 ,489 ,"Table S.1 of Evans et al., 2022",
  "Tule Lake", "Caspian Tern",2020 , 72, "Table S.1 of Evans et al., 2022",
  "Tule Lake", "American White Pelican", 2021 ,338 ,"Table 1 of Banet et al., 2024",
  "Tule Lake", "Double-crested Cormorant", 2021,326 ,"Table 1 of Banet et al., 2024",
  "Tule Lake", "Caspian Tern", 2021, 0,"Table 1 of Banet et al., 2024",
  "Tule Lake", "Gulls", 2021,NA ,"Table 1 of Banet et al., 2024",
  "Tule Lake", "American White Pelican", 2024 ,73 ,"Table 1 of Banet et al., 2025",
  "Tule Lake", "Double-crested Cormorant", 2024,2044 ,"Table 1 of Banet et al., 2025"
  )

```


```{r}
piscivorous_waterbirds |> 
  filter(!is.na(count)) |> 
  ggplot(aes(x = year, y = count, color = common_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ nesting_location, scales = "free_y") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Piscivorous Waterbird Counts Over Time",
    x = "Year",
    y = "Nest count",
    color = "Species"
  )

piscivorous_waterbirds |> 
  filter(!is.na(count)) |> 
  ggplot(aes(x = common_name, y = count, fill = common_name)) +
  geom_boxplot(alpha = 0.7) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Distribution of Nest Counts by Species",
    x = NULL,
    y = "Count"
  ) +
  theme(legend.position = "none")



```
##### Compare RTR/USGS data to eBird 


Aggregate piscivorous birds from eBird 
```{r}
pisci_birds <- c( "American White Pelican","Double-crested Cormorant", "Caspian Tern",  'Great Blue Herons', 'Great Egrets','Black-crowned Night Heron', 'California Gull', 'Ring-billed Gull')

ebd_pisci <- ebd_trunc |> 
  filter(common_name %in% pisci_birds) |> 
  filter(year(observation_date) %in% c(2009:2024)) |> 
  mutate(count = ifelse(observation_count == "X", 1, observation_count)) |> # if X mark it as 1
  mutate(count = as.numeric(count)) |> 
  select(-observation_count)

gulls <- ebd_pisci |> 
  filter(common_name %in% c('California Gull', 'Ring-billed Gull')) |> 
  group_by(year(observation_date)) |> 
  summarise(count = sum(count)) |> 
  mutate(common_name = "Gulls")

herons <- ebd_pisci |> 
  filter(common_name %in% c('Great Blue Herons', 'Great Egrets','Black-crowned Night Heron')) |> 
  group_by(year(observation_date)) |> 
  summarise(count = sum(count)) |> 
  mutate(common_name = "Herons")

rest_of_birds <- ebd_pisci |> 
  filter(common_name %in% c("American White Pelican","Double-crested Cormorant", "Caspian Tern")) |>
  group_by(common_name, year(observation_date)) |> 
  summarise(count = sum(count)) 

all_ebd_summary <- rest_of_birds |> 
  bind_rows(herons) |> 
  bind_rows(gulls) |> 
  rename(year = `year(observation_date)`) |> 
  mutate(dataset = "eBird")
```

Compare

```{r}
rtr_usgs_summary <- piscivorous_waterbirds |> 
  group_by(year, common_name) |> 
  summarise(count = sum(count)) |> 
  mutate(dataset = "RTR/USGS") 

bird_compare <- bind_rows(rtr_usgs_summary, all_ebd_summary)


ggplot(bird_compare, aes(x = year, y = count, color = dataset)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ common_name, scales = "free_y") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Comparison of Waterbird Counts: RTR/USGS vs eBird",
    x = "Year",
    y = "Count",
    color = "Dataset"
  )

```

Compare relative change; since eBird counts are much larger 

```{r}
bird_compare_index <- bird_compare |> 
  group_by(common_name, dataset) |> 
  mutate(index = count / mean(count, na.rm = TRUE)) |> 
  ungroup()

ggplot(bird_compare_index, aes(x = year, y = index, color = dataset)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ common_name) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Relative Trends in Waterbird Counts (Indexed)",
    x = "Year",
    y = "Index (mean = 1)",
    color = "Dataset"
  )

```

### Intermountain Shorebirds

https://www.imwss.org/

**Source**
-   Table 1: https://static1.squarespace.com/static/66b5066527e47020d0e0b7b0/t/673d0bf4ca1419202f3e9a2e/1732054007209/Intermountain+West+Shorebirds+-+Preliminary+Data+Summary+-+November+2024+Updated.pdf

```{r}
shorebirds <-  tribble(
  ~location,     ~common_name,          ~year,   ~season, ~count, 

  "Klamath Basin, CA and Oregon", "Shorebirds",  2022, "Fall", 10254,
  "Klamath Basin, CA and Oregon", "Shorebirds",  2023, "Spring", 155,
  "Klamath Basin, CA and Oregon", "Shorebirds",  2023, "Fall", 1678,
  "Klamath Basin, CA and Oregon", "Shorebirds",  2024, "Spring", 7297,
  "Klamath Basin, CA and Oregon", "Shorebirds",  2024, "Fall", 9625)
  
  

```


## Bird Groupings {#bird-groupings}

### Waterbirds

Grouping birds identified as part of the waterbird guild - including shorebirds, diving ducks, dabbling ducks, and other waterbirds. I've included primary habitat as well because the waterbirds are associated with multiple.

```{r}
waterbird_guild <- ebd_birdbase_with_waterbirds |> 
  filter(!is.na(waterbird_niche)) |> 
  group_by(waterbird_niche, primary_habitat, year = year(observation_date), month = month(observation_date)) |> 
  summarize(n_species = length(unique(common_name))) |> glimpse()

```

```{r}
waterbird_guild |> 
  ggplot(aes(x = as.factor(month), y = factor(year), fill = n_species)) +
  geom_tile(color = "white") +
  facet_wrap(~ waterbird_niche + primary_habitat) +
  scale_fill_viridis_c(name = "n_species") +
  labs(
    x = "Month",
    y = "Year",
    title = "Species Richness by Month and Year Across Waterbird and Primary Habitat Guilds"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines")
  )
```

```{r eval=FALSE, include=FALSE}

waterbird_guild |> 
  ggplot(aes(x = as.factor(month), y = n_species, group = year, fill = factor(year))) +
  geom_col(alpha = 0.6, stat = "identity", position = "dodge") +
  facet_wrap(~ wetland_niche, scales = "free_y") +
  labs(
    x = "Month",
    y = "Number of Species",
    color = "Year",
    title = "Monthly Species Richness by Waterbird Guild"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(1, "lines")
  )

```

### Habitat

Grouping the Klamath bird species by the primary habitat type. Habitat types include: `r paste0(unique(ebd_birdbase_with_waterbirds$primary_habitat), collapse = ", ")`. The number of species across months and years are grouped by each habitat type to create the `habitat` dataset.

```{r}
habitat <- ebd_birdbase_with_waterbirds |> 
  group_by(primary_habitat, year = year(observation_date), month = month(observation_date)) |> 
  summarize(n_species = length(unique(common_name))) |> 
  ungroup() |> 
  glimpse()

```

```{r}
habitat |> 
  ggplot(aes(x = as.factor(month), y = factor(year), fill = n_species)) +
  geom_tile(color = "white") +
  facet_wrap(~ primary_habitat) +
  scale_fill_viridis_c(name = "n_species") +
  labs(
    x = "Month",
    y = "Year",
    title = "Species Richness by Month and Year Across Habitat Guilds"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines")
  )
```

```{r eval=FALSE, include=FALSE}

habitat |> 
  ggplot(aes(x = as.factor(month), y = n_species, group = year, color = factor(year))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~ primary_habitat, scales = "free_y") +
  labs(
    x = "Month",
    y = "Number of Species",
    color = "Year",
    title = "Monthly Species Richness by Habitat Guild"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(1, "lines")
  )

```

### Foraging

Grouping the Klamath bird species by the primary diet type. Diet types include: `r paste0(unique(ebd_birdbase_with_waterbirds$primary_diet), collapse = ", ")`. The number of species across months and years are grouped by each diet type to create the `foraging` dataset.

```{r}
foraging <- ebd_birdbase_with_waterbirds |> 
  group_by(primary_diet, year = year(observation_date), month = month(observation_date)) |> 
  summarize(n_species = length(unique(common_name))) |> 
  ungroup() |> 
  glimpse()

```

```{r}
foraging |> 
  ggplot(aes(x = as.factor(month), y = factor(year), fill = n_species)) +
  geom_tile(color = "white") +
  facet_wrap(~ primary_diet) +
  scale_fill_viridis_c(name = "n_species") +
  labs(
    x = "Month",
    y = "Year",
    title = "Species Richness by Month and Year Across Foraging Guilds"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines")
  )
```

```{r eval=FALSE, include=FALSE}

foraging |> 
  ggplot(aes(x = as.factor(month), y = n_species, group = year, color = factor(year))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~ primary_diet, scales = "free_y") +
  labs(
    x = "Month",
    y = "Number of Species",
    color = "Year",
    title = "Monthly Species Richness by Foraging Guild"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(1, "lines")
  )

```

### Build Guild based on diet and habitat

Bird species were grouped into functional guilds based on a combination of primary diet and primary habitat. Diet × habitat guilds were defined for each observation by pairing the species’ primary diet category with its primary habitat category. Species richness was summarized by guild across months and years, and guilds were ranked by total species richness.

```{r}
habitat_diet_guild <- ebd_birdbase_with_waterbirds |> 
  mutate(guild = paste(primary_diet, primary_habitat, sep = " – ")) |> 
  group_by(year = year(observation_date), month = month(observation_date), primary_diet, primary_habitat, waterbird_niche, guild) |> 
  summarize(n_species = length(unique(common_name))) |> 
  ungroup() 

```

There are `r length(unique(habitat_diet_guild$guild))` guilds. The top 10 are:

```{r}
habitat_diet_guild |> 
  group_by(guild) |> 
  summarise(
    total_species = sum(n_species),
    .groups = "drop"
  ) |> 
  slice_max(total_species, n = 10) |> 
  ggplot(
    aes(x = fct_reorder(guild, total_species),
        y = total_species)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Top 10 diet × habitat guilds by species richness",
    x = "Guild",
    y = "Total number of species"
  ) +
  theme_minimal()

```

```{r}
habitat_diet_guild |> 
  group_by(guild) |> 
  summarise(total_species = sum(n_species)) |> 
  arrange(desc(total_species)) |> 
  ggplot(aes(x = fct_reorder(guild, total_species), y = total_species)) +
  geom_col() + 
  labs(
    x = "Guild",
    y = "Total number of species"
  ) + 
  coord_flip() 

```

```{r}
top_guilds <- habitat_diet_guild |> 
  group_by(guild) |> 
  summarise(
    total_species = sum(n_species),
    .groups = "drop"
  ) |> 
  slice_max(total_species, n = 10) 
top10_names <- top_guilds$guild

habitat_diet_guild |>
  mutate(
    guild = paste(primary_diet, primary_habitat, sep = " – "),
    month = factor(month, levels = 1:12, labels = month.abb)
  ) |>
  filter(guild %in% top10_names) |> 
  group_by(month, guild) |>
  summarise(n_species = sum(n_species), .groups = "drop") |> 
  ggplot(
    aes(x = month, y = n_species, group = guild, color = guild)) +
  geom_line() +
  labs(
    title = "Seasonal dynamics of the top 10 diet × habitat guilds",
    x = "Month",
    y = "Number of species",
    color = "Guild"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```

## This is the final *draft* dataset:

Bird species richness can be grouped by their primary habitat, diet or a combination of the two (`guild`). Also included is the waterbird niche, which can be used when applicable. Data is provided for the number of species from 2015 to 2025 for every month.

We are happy to provide a more granular representation of `guild` with suggested input on grouping.

Data can be accessed as `klamathWildlifeData::klamath_birds`

```{r}

klamath_birds <- habitat_diet_guild
klamath_birds |> glimpse()

usethis::use_data(klamath_birds, overwrite = TRUE)

```

```{r}
knitr::knit_exit()
```

## Other figures

```{r}

ggplot(habitat_diet_guild, aes(x = as.factor(month), y = primary_habitat, fill = n_species)) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_wrap(~ primary_diet, scales = "free_y") +
  scale_fill_viridis_c(name = "N species") +
  labs(
    title = "Species richness by habitat and month, faceted by primary diet",
    x = "Month",
    y = "Primary habitat"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}
habitat_diet_guild |> 
  ggplot(aes(
    x = primary_diet,
    y = primary_habitat,
    fill = n_species
  )) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(name = "N species") +
  labs(
    title = "Diet × Habitat species richness (guild candidates)",
    x = "Primary diet",
    y = "Primary habitat"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(1, "lines")
  )

```
