---
title: "Klamath Bird Data - Synthsized"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE, 
  fig.width = 12, 
  fig.height = 12
)

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(size = 11, face = "bold"),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14)
    )
)
```

### Build Dataset

1.  Pull in eBird data:

\*see: `ebird-data-pull.Rmd` \* for details on ebird data

```{r}
ebd_all <- readRDS(here::here("data", "ebd_klamath_combined.rds")) |> glimpse()
```

2.  Pull in BIRDBASE[^1]:

[^1]: Şekercioğlu, Çağan H; Kittelberger, Kyle D.; Mota, Flávia; Buxton, Amy N.; Orton, Nikolas; DeNiro, Adara; et al. (2025). BIRDBASE: A Global Database of Avian Biogeography, Conservation, Ecology and Life History Traits. figshare. Dataset. <https://doi.org/10.6084/m9.figshare.27051040.v1>

This dataset provides a thorough, literature based review of all bird species and includes habitat and foraging information that can be used to assign birds to guilds.

```{r}
birdbase <- readxl::read_excel(here::here('data-raw', 'BIRDBASE v2025.1 Sekercioglu et al. Final.xlsx'), sheet = "Data", skip = 1) |> 
  janitor::clean_names() |> glimpse()
```

3.  Merge ebird with BIRDBASE to get guild information:

The two datasets can easily be merged by the `scientific name`. The final dataset was heavily truncated to include on the data that is needed for this analysis.

```{r}
ebd_trunc <- ebd_all |> 
  select(common_name, scientific_name, observation_count, latitude, longitude, observation_date)

birdbase_trunc <- birdbase |> 
  select(scientific_name = e_bird_clements_v2024b, primary_diet, primary_habitat, species, genus, order)

ebd_birdbase <- ebd_trunc |> 
  left_join(birdbase_trunc) |> glimpse()
```

### Habitat Guild

Grouping the Klamath bird species by the primary habitat type. Habitat types include: `r paste0(unique(ebd_birdbase$primary_habitat), collapse = ", ")`. The number of species across months and years are grouped by each habitat type to create the `habitat_guild` dataset. 

```{r}
habitat_guild <- ebd_birdbase |> 
  group_by(primary_habitat, year = year(observation_date), month = month(observation_date)) |> 
  summarize(n_species = length(unique(common_name))) |> 
  ungroup() |> 
  glimpse()

```

```{r}
habitat_guild |> 
  ggplot(aes(x = as.factor(month), y = factor(year), fill = n_species)) +
  geom_tile(color = "white") +
  facet_wrap(~ primary_habitat) +
  scale_fill_viridis_c(name = "n_species") +
  labs(
    x = "Month",
    y = "Year",
    title = "Species Richness by Month and Year Across Habitat Guilds"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines")
  )
```

```{r}

habitat_guild |> 
  ggplot(aes(x = as.factor(month), y = n_species, group = year, color = factor(year))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~ primary_habitat, scales = "free_y") +
  labs(
    x = "Month",
    y = "Number of Species",
    color = "Year",
    title = "Monthly Species Richness by Habitat Guild"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(1, "lines")
  )

```

### Foraging Guild

Grouping the Klamath bird species by the primary diet type. Diet types include: `r paste0(unique(ebd_birdbase$primary_diet), collapse = ", ")`. The number of species across months and years are grouped by each diet type to create the `foraging_guild` dataset. 

```{r}
foraging_guild <- ebd_birdbase |> 
  group_by(primary_diet, year = year(observation_date), month = month(observation_date)) |> 
  summarize(n_species = length(unique(common_name))) |> 
  ungroup() |> 
  glimpse()

```

```{r}
foraging_guild |> 
  ggplot(aes(x = as.factor(month), y = factor(year), fill = n_species)) +
  geom_tile(color = "white") +
  facet_wrap(~ primary_diet) +
  scale_fill_viridis_c(name = "n_species") +
  labs(
    x = "Month",
    y = "Year",
    title = "Species Richness by Month and Year Across Foraging Guilds"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines")
  )
```

```{r}

foraging_guild |> 
  ggplot(aes(x = as.factor(month), y = n_species, group = year, color = factor(year))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~ primary_diet, scales = "free_y") +
  labs(
    x = "Month",
    y = "Number of Species",
    color = "Year",
    title = "Monthly Species Richness by Foraging Guild"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(1, "lines")
  )

```
