---
title: "Klamath Bird Data - Synthsized"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12, 
  fig.height = 12
)

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14)
    )
)
```

## Questions

1.  Do we want individual species tracked to guild?
2.  Should we add a geographic component to the dataset?
3,  Can we provide more speficity in guild structure?

## Build Dataset

### 1. Pull in eBird data:

\*see: `ebird-data-pull.Rmd` \* for details on ebird data. This data is filtererd to the Klamath Basin. 

```{r}
ebd_all <- readRDS(here::here("data", "ebd_klamath_combined.rds")) |> glimpse()
```

### 2. Pull in BIRDBASE[^1]:

[^1]: Şekercioğlu, Çağan H; Kittelberger, Kyle D.; Mota, Flávia; Buxton, Amy N.; Orton, Nikolas; DeNiro, Adara; et al. (2025). BIRDBASE: A Global Database of Avian Biogeography, Conservation, Ecology and Life History Traits. figshare. Dataset. <https://doi.org/10.6084/m9.figshare.27051040.v1>

This dataset provides a thorough, literature based review of all bird species and includes habitat and foraging information that can be used to assign birds to guilds.

The `primary habitat` and `primary diet` values were used but it's worth noting that BIRDBASE provides information on how each bird species fits into each type. This could help if a more involved analysis is requested.

```{r}
birdbase <- readxl::read_excel(here::here('data-raw', 'BIRDBASE v2025.1 Sekercioglu et al. Final.xlsx'), sheet = "Data", skip = 1) |> 
  janitor::clean_names() |> glimpse()
```

3.  Merge ebird with BIRDBASE to get guild information:

The two datasets can easily be merged by the `scientific name`. The final dataset was heavily truncated to include on the data that is needed for this analysis.

```{r}
ebd_trunc <- ebd_all |> 
  select(common_name, scientific_name, observation_count, latitude, longitude, observation_date)

birdbase_trunc <- birdbase |> 
  select(scientific_name = e_bird_clements_v2024b, primary_diet, primary_habitat, species, genus, order)

ebd_birdbase <- ebd_trunc |> 
  left_join(birdbase_trunc) 
```
### 3. Add waterbird niche

Based off of table 2 of Donnelly et al., 2022[^2]

[^2]: Donnelly, J. P., Moore, J. N., Casazza, M. L., & Coons, S. P. (2022). Functional wetland loss drives emerging risks to waterbird migration networks. *Frontiers in Ecology and Evolution, 10*, 844278. <https://doi.org/10.3389/fevo.2022.844278>

```{r}
waterbirds <- tribble(
  ~waterbird_niche,              ~common_name,
  # Shorebirds
  "Shorebirds",        "American Avocet",
  "Shorebirds",        "Black-necked Stilt",
  "Shorebirds",        "Dunlin",
  "Shorebirds",        "Greater Yellowlegs",
  "Shorebirds",        "Lesser Yellowlegs",
  "Shorebirds",        "Long-billed Dowitcher",
  "Shorebirds",        "Marbled Godwit",
  "Shorebirds",        "Western Sandpiper",
  "Shorebirds",        "Willet",
  "Shorebirds",        "Wilson's Phalarope",
  "Shorebirds",        "Wilson's Snipe",
  "Shorebirds",        "Whimbrel",
  
  # Dabbling ducks
  "Dabbling ducks",    "American Wigeon",
  "Dabbling ducks",    "Cinnamon Teal",
  "Dabbling ducks",    "Gadwall",
  "Dabbling ducks",    "Green-winged Teal",
  "Dabbling ducks",    "Mallard",
  "Dabbling ducks",    "Northern Pintail",
  "Dabbling ducks",    "Northern Shoveler",
  
  # Diving ducks
  "Diving ducks",      "Barrow's Goldeneye", 
  "Diving Ducks",      "Common Goldeneye",
  "Diving ducks",      "Bufflehead",
  "Diving ducks",      "Canvasback",
  "Diving ducks",      "Greater Scaup", 
  "Diving ducks",      "Lesser Scaup",
  "Diving ducks",      "Redhead",
  "Diving ducks",      "Ring-necked Duck",
  "Diving ducks",      "Ruddy Duck",
  
  # Other waterbirds
  "Other waterbirds",  "American Bittern",
  "Other waterbirds",  "American Coot",
  "Other waterbirds",  "Black Tern",
  "Other waterbirds",  "Eared Grebe",
  "Other waterbirds",  "Least Bittern",
  "Other waterbirds",  "Western Grebe",
  "Other waterbirds",  "White-faced Ibis"
) 

knitr::kable(waterbirds)

```
```{r}
ebd_birdbase_with_waterbirds <- ebd_birdbase |> 
  left_join(waterbirds) |> glimpse()
```

## Explore potential groupings

### Waterbirds 

Grouping birds identified as part of the waterbird guild - including shorebirds, diving ducks, dabbling ducks, and other waterbirds. I've included primary habitat as well because the waterbirds are associated with multiple. 

```{r}
waterbird_guild <- ebd_birdbase_with_waterbirds |> 
  filter(!is.na(waterbird_niche)) |> 
  group_by(waterbird_niche, primary_habitat, year = year(observation_date), month = month(observation_date)) |> 
  summarize(n_species = length(unique(common_name))) |> glimpse()

```

```{r}
waterbird_guild |> 
  ggplot(aes(x = as.factor(month), y = factor(year), fill = n_species)) +
  geom_tile(color = "white") +
  facet_wrap(~ waterbird_niche + primary_habitat) +
  scale_fill_viridis_c(name = "n_species") +
  labs(
    x = "Month",
    y = "Year",
    title = "Species Richness by Month and Year Across Waterbird and Primary Habitat Guilds"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines")
  )
```

```{r eval=FALSE, include=FALSE}

waterbird_guild |> 
  ggplot(aes(x = as.factor(month), y = n_species, group = year, fill = factor(year))) +
  geom_col(alpha = 0.6, stat = "identity", position = "dodge") +
  facet_wrap(~ wetland_niche, scales = "free_y") +
  labs(
    x = "Month",
    y = "Number of Species",
    color = "Year",
    title = "Monthly Species Richness by Waterbird Guild"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(1, "lines")
  )

```

### Habitat

Grouping the Klamath bird species by the primary habitat type. Habitat types include: `r paste0(unique(ebd_birdbase_with_waterbirds$primary_habitat), collapse = ", ")`. The number of species across months and years are grouped by each habitat type to create the `habitat` dataset.

```{r}
habitat <- ebd_birdbase_with_waterbirds |> 
  group_by(primary_habitat, year = year(observation_date), month = month(observation_date)) |> 
  summarize(n_species = length(unique(common_name))) |> 
  ungroup() |> 
  glimpse()

```

```{r}
habitat |> 
  ggplot(aes(x = as.factor(month), y = factor(year), fill = n_species)) +
  geom_tile(color = "white") +
  facet_wrap(~ primary_habitat) +
  scale_fill_viridis_c(name = "n_species") +
  labs(
    x = "Month",
    y = "Year",
    title = "Species Richness by Month and Year Across Habitat Guilds"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines")
  )
```

```{r eval=FALSE, include=FALSE}

habitat |> 
  ggplot(aes(x = as.factor(month), y = n_species, group = year, color = factor(year))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~ primary_habitat, scales = "free_y") +
  labs(
    x = "Month",
    y = "Number of Species",
    color = "Year",
    title = "Monthly Species Richness by Habitat Guild"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(1, "lines")
  )

```

### Foraging

Grouping the Klamath bird species by the primary diet type. Diet types include: `r paste0(unique(ebd_birdbase_with_waterbirds$primary_diet), collapse = ", ")`. The number of species across months and years are grouped by each diet type to create the `foraging` dataset.

```{r}
foraging <- ebd_birdbase_with_waterbirds |> 
  group_by(primary_diet, year = year(observation_date), month = month(observation_date)) |> 
  summarize(n_species = length(unique(common_name))) |> 
  ungroup() |> 
  glimpse()

```

```{r}
foraging |> 
  ggplot(aes(x = as.factor(month), y = factor(year), fill = n_species)) +
  geom_tile(color = "white") +
  facet_wrap(~ primary_diet) +
  scale_fill_viridis_c(name = "n_species") +
  labs(
    x = "Month",
    y = "Year",
    title = "Species Richness by Month and Year Across Foraging Guilds"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines")
  )
```

```{r eval=FALSE, include=FALSE}

foraging |> 
  ggplot(aes(x = as.factor(month), y = n_species, group = year, color = factor(year))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~ primary_diet, scales = "free_y") +
  labs(
    x = "Month",
    y = "Number of Species",
    color = "Year",
    title = "Monthly Species Richness by Foraging Guild"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(1, "lines")
  )

```

### Build Guild based on diet and habitat

Bird species were grouped into functional guilds based on a combination of primary diet and primary habitat. Diet × habitat guilds were defined for each observation by pairing the species’ primary diet category with its primary habitat category. Species richness was summarized by guild across months and years, and guilds were ranked by total species richness.

```{r}
habitat_diet_guild <- ebd_birdbase_with_waterbirds |> 
  mutate(guild = paste(primary_diet, primary_habitat, sep = " – ")) |> 
  group_by(year = year(observation_date), month = month(observation_date), primary_diet, primary_habitat, waterbird_niche, guild) |> 
  summarize(n_species = length(unique(common_name))) |> 
  ungroup() 

```

There are `r length(unique(habitat_diet_guild$guild))` guilds. The top 10 are:

```{r}
habitat_diet_guild |> 
  group_by(guild) |> 
  summarise(
    total_species = sum(n_species),
    .groups = "drop"
  ) |> 
  slice_max(total_species, n = 10) |> 
  ggplot(
    aes(x = fct_reorder(guild, total_species),
        y = total_species)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Top 10 diet × habitat guilds by species richness",
    x = "Guild",
    y = "Total number of species"
  ) +
  theme_minimal()

```

```{r}
habitat_diet_guild |> 
  group_by(guild) |> 
  summarise(total_species = sum(n_species)) |> 
  arrange(desc(total_species)) |> 
  ggplot(aes(x = fct_reorder(guild, total_species), y = total_species)) +
  geom_col() + 
  labs(
    x = "Guild",
    y = "Total number of species"
  ) + 
  coord_flip() 

```

```{r}
top_guilds <- habitat_diet_guild |> 
  group_by(guild) |> 
  summarise(
    total_species = sum(n_species),
    .groups = "drop"
  ) |> 
  slice_max(total_species, n = 10) 
top10_names <- top_guilds$guild

habitat_diet_guild |>
  mutate(
    guild = paste(primary_diet, primary_habitat, sep = " – "),
    month = factor(month, levels = 1:12, labels = month.abb)
  ) |>
  filter(guild %in% top10_names) |> 
  group_by(month, guild) |>
  summarise(n_species = sum(n_species), .groups = "drop") |> 
  ggplot(
    aes(x = month, y = n_species, group = guild, color = guild)) +
  geom_line() +
  labs(
    title = "Seasonal dynamics of the top 10 diet × habitat guilds",
    x = "Month",
    y = "Number of species",
    color = "Guild"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```

## This is the final dataset:

Bird species richness can be grouped by their primary habitat, diet or a combination of the two (`guild`). Also included is the waterbird niche, which can be used when applicable. Data is provided for the number of species from 2015 to 2025 for every month. 

We are happy to provide a more granular representation of `guild` with suggested input on grouping.

```{r}

klamath_birds <- habitat_diet_guild
klamath_birds |> glimpse()

usethis::use_data(klamath_birds)

```

```{r}
knitr::knit_exit()
```

## Other figures

```{r}

ggplot(habitat_diet_guild, aes(x = as.factor(month), y = primary_habitat, fill = n_species)) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_wrap(~ primary_diet, scales = "free_y") +
  scale_fill_viridis_c(name = "N species") +
  labs(
    title = "Species richness by habitat and month, faceted by primary diet",
    x = "Month",
    y = "Primary habitat"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}
habitat_diet_guild |> 
  ggplot(aes(
    x = primary_diet,
    y = primary_habitat,
    fill = n_species
  )) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(name = "N species") +
  labs(
    title = "Diet × Habitat species richness (guild candidates)",
    x = "Primary diet",
    y = "Primary habitat"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(1, "lines")
  )

```
